---
title: "Liver Disease in Indian Patients"
author: "Waseem Medhat"
output:
  pdf_document:
    df_print: kable
    includes:
      in_header: preamble.tex
    number_sections: yes
  word_document: default
  html_document:
    df_print: kable
    number_sections: yes
subtitle: Exploratory Data Analysis Project
urlcolor: blue
---


```{r init, echo=FALSE}
# read in the reference script
knitr::read_chunk("analysis_script.r")

# set chunk options
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9
)

# NOTE: The chunks below are empty as they refer to the code in the script read
#       above. This was made for maximum reproducibility and coherence between
#       interactive analysis done in the script and report generation instead of
#       copying/pasting.
```

```{r packages}
```

```{r data}
```


This project demonstrates the use of exploratory data analysis (EDA) to search
for interesting information or insights within data. Data visualization and
statistical summaries are the main tools in EDA, and together they can be very
powerful in data analysis.

The data set used in this project contains data on liver disease in Indian
patients^[Data source: [Kaggle](https://www.kaggle.com/uciml/indian-liver-patient-records)].
It is a relatively small data set with `r dim(hep)[2]` attributes (columns) on
`r dim(hep)[1]` patients (rows). The outcome variable is a binary categorical
one indicating whether or not a particular case has liver disease, consequently
this will be the center around which the exploration will be done. In other
words, even when other relationships are considered, the primary goal here is
finding any patterns or associations of liver disease with the other variables,
which include:

+ age,
+ gender,
+ total bilirubin (in mg/dL),
+ direct, i.e. conjugated, bilirubin (in mg/dL),
+ alkaline phosphatase (ALP, in IU/L),
+ alanine aminotransferase (ALT, in IU/L),
+ aspartate aminotransferase (AST, in IU/L),
+ total proteins (in g/dL),
+ albumin (in g/dL), and
+ albumin/globulin (A/G) ratio.

There are usually preliminary data preparation/cleaning steps that should be
done first. For this data set:

+ There are spelling mistakes in the column names `Total_Protiens`,
  `Alamine_Aminotransferase`, and `Alkaline_Phosphotase`, which were corrected
  to `Total_Proteins`, `Alanine_Aminotransferase` and `Alkaline_Phosphatase`,
  respectively.
+ For the sake of clarity, a new column named `Diseased` was created from the
  `Dataset` variable, which corresponds to the disease categories, with its
  levels being "Yes" for diseased cases and "No" for healthy ones. The original
  variable was removed from the data set.
+ Other derived columns were created in the analysis process as transformations
  (base 10 log) of skewed variables to be able to analyze them.
  
***

# Univariate Analysis

## Data Types

```{r datatypes}
```

Most of our variables are numeric, and they represent various blood
concentrations of a particular compound or enzyme along with an age variable. 
There are two remaining categorical variables: disease category and gender, both
of which are binary.

## Missing Values

We can start by analyzing each variable on its own to take a closer look at the
structure of the data set. But before starting the analysis, we should check the
data set for missing values.

```{r missing}
```

There are no missing data in all variables except A/G ratio, and even the
missing values in that variable constitute
`r percent_missing["Albumin_and_Globulin_Ratio"]`% of the data
(`r number_missing["Albumin_and_Globulin_Ratio"]` cases). Therefore, it is safe
to use all variables in the analysis and drop the missing values when necessary
without affecting our analysis.

But it is still better to think about why these values were missing. One
possible explanation comes from the fact that it is a ratio of albumin to
globulin levels. We do not have the globulin data, but they could have been
missing which resulted in A/G ratio being missing as well.

## Numeric Variables

Histograms are a good method to examine the distributions of numeric variables.

```{r uni_histograms}
```

It seems that age, albumin, and total proteins have roughly symmetric
distributions. A/G ratio has a few outliers. These will be worth looking into
further in the analysis.

```{r uni_ag_outliers}
```

The rest of the variables, which are either bilirubin levels or liver enzymes,
are heavily right skewed. We can recreate the plots of these variables with a
log-transformed X axis to have a better look at their distribution.

```{r uni_log_histograms}
```

```{r uni_bili_tables}
```

Even with a log transformation, the histograms still show long tails. Also,
total and direct bilirubin distributions look peculiar: each has a very high
spike at a single bin which corresponds to a value of `r names(btot_table)[1]`
mg/dL for total bilirubin and `r names(bdir_table)[1]` mg/dL for direct
bilirubin. The histogram of direct bilirubin shows some discreteness in the
lower bin values which is only a result of the data being recorded/rounded with
one decimal precision.

```{r uni_bdir_below_1}
```

Before leaving the histograms, it might be a good idea to decrease the binwidth
as it might uncover some hidden details.

```{r uni_histograms_more_bins}
```

We can quickly notice the discreteness in the distribution of A/G ratio which
results from the same cause as bilirubin levels. Moreover, for the distribution
of albumin, the same peaks can be observed before and after decreasing the
binwidth, which means that the most common albumin levels do not lie in ranges
but specific values. Examining the exact numbers will give us a better idea.

```{r uni_albumin_table}
```

The most common values are `r names(albumin_table)[1]` and
`r names(albumin_table)[2]` mg/dL. Even the next most common values lie near 
those values.

Now, we take a closer look at the distribution of ages by using a bin for each
single year (i.e. a binwidth of 1).

```{r uni_age_histogram}
```

Interestingly, ages that are multiples of 5 are much more common than the
adjacent ones. This pattern can be found starting from 40 years. Could it be
that some old people report their ages rounded to the nearest 5? It might seem a
trivial topic to investigate compared to liver disease, but it is interesting
nonetheless.

The next plot zooms in on the last one and adds color emphasis on multiples of 5
to make it easier to look at the pattern.

```{r uni_age_histogram_zoom}
```

## Categorical Variables

The data set has only two categorical variables: gender and disease state. We can
look at frequencies and proportions of their categories using numerical
summaries and bar charts.

```{r uni_bars, fig.height=2}
```

```{r uni_freq}
```

Looking at the gender data, there are more males than females. Similarly,
there are more liver patients than healthy people.

## Univariate Analysis Insights {-}

This section serves as a reflection that connects findings in the data with
their actual implications. We use what we have found to build some intuitions to
pave the road for the next phase in the EDA process.

We saw that bilirubin and liver enzyme data have almost the same distribution
pattern. This is due to a property they all have in common: their blood levels
are very low in people with healthy livers but increase with liver disease.
Bilirubin levels are kept low by the liver as it excreted by the liver, while
the enzymes are normally present inside the liver cells. Either way, these
compounds accumulate in the blood with liver damage and show high levels in lab
tests^[http://www.clevelandclinicmeded.com/medicalpubs/diseasemanagement/hepatology/guide-to-common-liver-tests/].
Because there is no upper limit to these values, their frequencies are scattered
across a very broad range, while normal values are condensed in a very narrow
range, causing these skewed distributions even though we have many more patients
than healthy people. Another thing to be noted is that the liver is the main
source of albumin^[https://en.wikipedia.org/wiki/Albumin#Serum_albumin], but
this is not the case for globulin (not every type of globulin is produced in the
liver)^[https://en.wikipedia.org/wiki/Globulin]. This implies that low albumin
levels, and consequently A/G ratio, correspond to liver damage.

***

# Bivariate Analysis

After using univariate analysis to get a sense of the structure of the data set
and the information contained within each variable, it is time to dig deeper
into the relationships between variables. As stated earlier, the main focus is
on the disease state as an outcome variable, but other relationships between
explanatory variables will be examined as well.

## Correlations

A correlation matrix can be a very concise way to look at relationships between
numeric variables.

```{r bi_correlations, fig.height=6}
```

Some unsurprisingly strong correlations are direct bilirubin with total
bilirubin (the former is only a subset of the latter), AST with ALT (both are
liver enzymes), and albumin (being the most abundant protein) with total
proteins.

```{r ag_cor_alb}
```

The correlation between albumin and A/G ratio is `r round(ag_cor_alb, 2)` which
is also high but not as high as other strong correlations because globulin
introduces some variability in the ratio that is not shared with albumin.

Although the correlation matrix is informative, it is better to look at
scatterplots to examine the relationships in more detail or to ensure that no
non-linear trends were missed. (Note: because bilirubin and liver enzymes are
heavily skewed, they were log-transformed to clarify the scatterplots.)

```{r log_vars}
```

```{r bi_scatterplots, fig.height=6}
```

```{r ag_cor_bdir}
```


In this particular case, no important insights can be gleaned from the
scatterplots that the correlation coefficients did not show. However, there is
something to notice: in the A/G ratio vs. (log) direct bilirubin, the outliers
in A/G ratio lie in a position that would increase the slope of the fitted line
(i.e. result in a more positive correlation). As a result, the correlation
coefficient of `r round(ag_cor_bdir, 2)` between A/G ratio and direct bilirubin
could be biased by the outliers.

Here is a single scatterplot for clarification. (Note: the points were jittered
to avoid overplotting.)

```{r bi_ag_bdir_scatterplot}
```

We can see a slightly negative trend. Higher values of direct bilirubin are
associated, even if weakly, with lower A/G ratio. This makes sense since low
A/G ratio and high direct bilirubin are both indicators of liver disease.

If we recalculated the correlation coefficient between these 2 variables after
removing the outliers, we would get a value of `r round(ag_cor_bdir_noout, 2)`,
which indicates a stronger correlation (to the negative side). It is important,
however, to carefully examine those outliers before deciding to remove them.
Outliers are not necessarily erroneous; maybe they are a special group of
observations that should be analyzed "in isolation" instead of completely
dismissing them.

## Gender vs. Disease

We might want to know if liver disease is more common in males or females. Also,
in later steps we may analyze variables that are grouped by both gender and
disease state. So we might first examine the distribution of genders across the
two disease states.

```{r bi_bars, fig.height=3}
```

The proportion of males is not substantially different across the two states.
Both roughly reside near 0.75. We can also look at counts using a crosstab.

```{r crosstab}
```

## Numeric Variables vs. Disease

Here we examine age and lab tests and see how their values differ between people
who have liver disease and those who do not. This is where we test if the
intuitions developed in the univariate analysis phase are correct.

```{r bi_boxplots}
```

+ **Age:** Liver disease patients are typically older than healthy people. This
  can be mainly due to the fact that liver diseases are chronic.
+ **Albumin:** Since, as mentioned above, albumin is produced by the liver, a
  decrease in its levels is expected to be associated with the presence of liver
  disease.
+ **A/G ratio:** The same concept with albumin applies here.
  However, the outliers spotted earlier seem to represent liver disease
  patients. This does not align with our intuition or the rest of the data.
+ **Total proteins:** There is only a slight change in total proteins between 
  diseased and healthy people.
+ **Bilirubin and liver enzymes:** It is clear that their values stretch to a
  high range in diseased patients. But in the case of ALP, there is one outlier
  in the healthy people with an unusually high value.

```{r bi_log_boxplots}
```

Since boxplots do not show the means, we can separately summarize them into a
table.

```{r bi_means}
```

## Age vs. Gender

We found in the univariate analysis that ages that are multiples of 5 are
particularly common. We can further refine this by splitting the distribution
across genders and see how they differ.

```{r bi_age_gender_facet}
```

Zooming in...

```{r bi_age_gender_zoom}
```

We can see that the pattern we found in the univariate histogram is more visible
in males than in females.

## Bivariate Analysis Insights {-}

Through bivariate analysis, we started understanding relationships between
variables in the data set and testing our intuitions as we found with the
values of blood tests. We also refined findings of univariate analysis when we
found that the age pattern is more visible in 40+ year old males.

We also discovered an interesting outlier in which a non-diseased case had high
ALP level. The most likely explanation is that we do not know if the cases that
have no liver disease are actually free from other conditions, and ALP level is
not associated only with liver
diseases^[https://www.healthline.com/health/alp#uses]. Besides the new outlier,
we found new information about the previously discovered ones. Similar to the
ALP outlier, there were 3 unexpectedly high A/G ratios in patients with no liver
disease. Again these unusual values could be attributed to other
conditions^[https://www.idexx.eu/globalassets/documents/parameters/8090-us-agratio-interpretive-summary.pdf].

_A word of caution_: all theses speculations assume that there were no errors in
measurement or data entry. In situations where it is feasible to consult those
responsible for data collection, it would be advisable to do so.

***

# Multivariate Analysis

At this point of the EDA process, important relationships and features of the
data have been discovered. Next, we explore more complex relationships and
interactions that involve more than two variables.

## Numeric Variables vs. Gender and Disease

The boxplots created earlier are split (using color) by gender, and scales were
log-transformed when necessary.

```{r multi_disease_gender, fig.height=8}
```

In most cases, non-diseased men and women are not substantially different.
However, diseased males have typically have lower levels of albumin and higher
levels of bilirubin and enzymes than diseased females. Also, it seems as if
there is no box in the direct bilirubin distribution for healthy females. This
indicates that the 3 quartiles (1st, median, and 3rd) are equal. We can back
this up by summary statistics.

```{r multi_bdir_female_no}
```

## Age vs. Gender and Disease

We can build up on the age vs. gender histograms by faceting by the disease
variable as well.

```{r multi_age}
```

Unfortunately, for a data set this small, we over-split the age variable.
Frequencies in the non-diseased cases are all below 10 which is too low to make
a solid judgment, so we cannot go beyond what we found in the bivariate
analysis.

## Final Look at Outliers

Previously we found outliers in A/G ratio and another one in ALP. The next plot
is a scatterplot of the two variables faceted by both gender and disease. This
will give us an idea if the outliers are related in any way. (This time ALP was
used without log-transformation because we are interested in the outliers.)

```{r multi_outliers}
```

The 3 A/G ratio outliers are males with liver disease and have seemingly typical
ALP values. The ALP outlier is a male with no liver disease and also has a
typical A/G value. This gives us no evidence that these outliers are related.

## Multivariate Analysis Insights {-}

We discovered that blood test results might differ between males and females
that do have liver disease. This could point out, without being a solid proof,
to a conclusion that males are affected by liver disease more severely than
females or that there is a confounder that is associated with gender (e.g.
physiological differences or behavioral risk factors) and cause this difference.

Other multivariate explorations were dead ends. The data set was too small to
explore age across two factors, and the suspicion that different outliers are
related turned out to be false.

***

# Final Plots and Summary

In this section the paradigm switches from exploratory, where we search for
information, to explanatory, where we present information. This switch will have
an effect on the structure and form of the plots we create which will be the
final presentables.
  
## First Plot

This plot shows the mean ALT level in patients vs. healthy people. It summarizes
the findings we found in the boxplots of liver enzymes grouped by disease
category.

```{r final_1_summary}
```

```{r final_1, fig.height=3}
```

The following points are the choices made in the plot with and the reasons
behind those choices:

+ Instead of boxplots, like the exploratory plots, bar charts are easier to look
  at and interpret as it is known by a larger number of people from different
  domains and levels of expertise.
+ Only a single plot of a single enzyme was used. Since the plot is meant to
  deliver a single piece of information, using multiple plots for each enzyme
  is redundant.
+ There are a title and a subtitle in order to make the plot informative on its
  own without having to refer to any text. The title shows the main conclusion,
  and the subtitle tells the finding which led to that conclusion.
+ Instead of putting the disease categories as Yes/No, names were changed to
  "Diseased" and "Healthy" which is a slight addition to both overall polish and
  ease of interpretation.
+ Error bars were added to remedy the loss of information that is due to
  switching from a boxplot to a bar chart. A caption was added to inform that
  the error bars represent 95% confidence intervals.


## Second Plot

The next plot shows how in liver disease the mean ALT increases more
significantly in males than females.

```{r final_2_summary}
```

```{r final_2}
```

For this plot:

+ Most points considered in the first plot apply here, including the choice of
  a bar chart, adding error bars, a title, a subtitle, and a caption.
+ The color scale now encodes gender, and a legend was added accordingly.


## Third Plot

The final plot shows the distribution of ages split across gender, same as the
one explored earlier, but refined visually.

```{r final_3}
```

For this plot:

+ A title and a subtitle were added.
+ The same color scale is used for consistency. But in this plot, it has an
  important contribution to the plot's effectiveness: the blue color is much
  stronger than yellow, and this makes the blue bars look more prominent at the
  first glance because those are were the emphasis should lie.
+ Breaks of 5 were used in the exploratory phase and kept here as they are.
+ Although the main question of the plot involves males only, the female part
  of the plot was kept to show that the pattern is less pronounced there.
+ There is one tiny detail: switching the Male/Female labels from the right to
  the left. I found it being on the right makes it slightly less visible "in
  the spotlight." Usually it is the left (axis) position that draws the
  attention first, and the frequency axis is not critically important in this 
  particular case, so putting the gender labels on the left is a better choice
  by that logic.
  
***

# Reflection

EDA has an iterative, brainstorming-like quality to it, and it does not always
have strict directions. One question can be posed based on ideas already formed
before analysis, but another can be posed after noticing a pattern or an
outlier. This means that the project by no means provides an exhaustive list of
available approaches in EDA for a data set, not even for _this_ data set, but
it is merely an example of few thought processes to follow in order to draw
information from data.

Most of the conclusions that we drew from analyzing this data set confirm (or
are confirmed by) available research and other resources. The age pattern, on
the other hand, is an example of a peculiar finding that comes from data
analysis but could be just a myth where one draws a line around a random set of
points and call it a shape. This is why finding and collecting more data will be
an improvement. First, such patterns need to be confirmed in multiple sets for
the same (Indian) or different populations. Also, more data will allow us to
zoom in on groups and facet by multiple factors.

Finally, since it has the word "exploratory" in its name, EDA is merely a tool
that introduces the analyst to the nature of the data. There are stronger
statistical procedures like inference and modeling that can help us make better
use and draw more solid conclusions from data.